### 2014-11-30 text/Chapter06.tex
Finish Ch6 and tweaks everywhere (Chapter 6)

\begin{itemize}
\item Structural coeffect calculus contains explicit rules for context manipulation
  (weakening, contraction, exchange). In flat coeffect calculus, these rules are not defined
  explicitly, but are admissible.

\item In structural coeffect calculus, the variable context is treated as a vector
  and is annotated with a vector of (scalar) coeffects. In flat coeffect calculus,
  the variable context is a set and is annotated with a single (scalar) coeffect.

\item In flat coeffect calculus, we distinguish between splitting of the context requirements
  and merging of context requirements ($\cpar$ and $\czip$, respectively). In structural
  coeffect calculus, the operations (which model splitting and appending vectors) are invertible 
  and so the structural coeffect algebra requires just $\atimes$.
\end{itemize}

@@

\subsection{Structure of coeffects}
In structural coeffect calculus, the structure annotation was formed by a vector of coeffect
sclars. The unified coeffect calculus is similar, but a \emph{vector} is replaced with a 
\emph{container}. The primitive coeffect annotations in the unified calculus are formed by
a \emph{coeffect scalar}, which remains the same as in structural coeffect calculus
(Definition~\ref{def:structural-scalar}). In this section, we refer to it as
\emph{\cclrd{unified coeffect scalar}} (and we repeat the definition below). Then we
define \emph{\sclrd{unified coeffect containers}} which determines how coeffect scalar values 
are attached to the free-variable context. Finally, we define the \emph{\aclrd{unified coeffect 
algebra}} which consists of shape-indexed coeffect scalar values.

@@

\begin{example}
Structural coeffect shape is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
where $\SH=\mathbb{N}$ and $\SP \sclrd{n}=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the 
number of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$, 
respectively, and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} = 
{|}\Gamma_1{|} + {|}\Gamma_2{|}$.

@@

\begin{example}
Flat coeffect shape is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP \sflat = \{ 0 \}$
with a constant function ${|}\Gamma{|}=\sflat$ and a trivial operation $\sflat\,\diamond\,\sflat = \sflat$.

@@

\begin{definition}
Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a 
\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \emph{\aclrd{unified 
coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero$ is a 
coeffect annotation for the empty context and $\azip, \apar$ are families of operations that 
combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
%
\begin{equation*}
\apar_{\sclrd{m},\sclrd{n}}, \azip_{\sclrd{m},\sclrd{n}} ~:~ \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
\end{equation*}
\end{definition}

\noindent
A coeffect algebra induces the following three additional operations:
%
\begin{equation*}
\begin{array}{rclcrcl}
 \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP \sunit}  \\
 \alift{\cclrd{x}} &\narrow{=}& \lambda \sunit . \cclrd{x}   \\
\\[-0.5em]
 \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP \sclrd{m}} \rightarrow \C^{\SP \sclrd{m}}   \\
 \cclrd{r}\, \aseq\, \aclrd{\textbf{s}} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\, \cseq\, \aclrd{\textbf{s}}(\sclrd{s}) \\
\\[-0.5em]
 \alen{-} &\narrow{:}& \C^{\SP \sclrd{m}} \rightarrow \sclrd{m}\\
 \alen{\aclrd{\textbf{r}}} &\narrow{=}& \sclrd{m}
\end{array}
\end{equation*}
%
The $\alift{-}$ operation lifts a scalar coeffect to a shape-indexed coeffect that is indexed by 
the singleton context shape. The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a 
vector by scalar. As we always use bold font for vectors and ordinary font for scalars (as well 
as distinct colour), using the same symbol is not ambiguous. We also tend to omit the subscript
$\sclrd{m}$ and write $\aseq$.

@@

\vspace{1em}
{\small b.) Structural rules for context manipulation:}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s'}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{s}} \apar \aclrd{\textbf{q}}} \vdash e : \tau}
\quad
(\cclrd{s'} \cleq \cclrd{s})
\end{equation*}
\begin{equation*}
\tyrule{weak}
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \apar \alift{ \czero }} \vdash e : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s}} \azip \alift{\cclrd{t}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{t}} \apar \alift{\cclrd{s}} \apar \aclrd{\textbf{q}}} \vdash e : \tau}
\quad
\begin{array}{l}
 \slen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
 \slen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
\end{array}
\end{equation*}
\begin{equation*}
\hspace{2.5em} 
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s}} \azip \alift{\cclrd{t}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \apar \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
~
\begin{array}{l}
 \slen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
 \slen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
\end{array}
\end{equation*}

\caption{Type system for the unified coeffect calculus}
\label{fig:unif-types}
\end{figure}

@@

\subsection{Unified coeffect type system}

The unified coeffect system in Figure~\ref{fig:unif-types} resembles the structural type system
shown in Figure~\ref{fig:struct-types}. Rather than explaining the rules one-by-one, we focus on
the key differences between the two.

The type system for the unified coeffect calculus is parameterized by a coeffect scalar 
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ together with a coeffect algebra $(\azip, \apar, \azero)$ 
and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$.

As in the structural system, free-variable contexts $\Gamma$ are treated as vectors modulo duplicate
use of variables -- the associativity is built-int. The order of variables matters, but can be changed 
using the structural rules. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}} \in \C^\SH$ 
are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
annotated with coeffects scalars. % $\cclrd{r}, \cclrd{s}, \cclrd{t} \in \C$.

\paragraph{Differences from structural system.}

% => \azero for constants 
% => \azip in assumptions (let, abs), \apar in conclusions (let, app)
% => the same also happened in structural rules
%      --> how does that make contraction etc. permissible?
% => <s,r> -> <s> x <r>

Constants (\emph{const}) and variable access (\emph{var}) annotate the context with special
values. Empty unused context is annotated with $\azero
\in \C^\sempty$, while singleton context 
is annotated with $\aunit \in \C^\sunit$.  Note that the shapes
$\sempty, \sunit$ match the shape of the variable contexts.

Lambda abstraction \emph{splits} the context requirements using $\azip$ into a coeffect 
$\aclrd{R}$ and a coeffect $\alift{\cclrd{s}}$ of a shape $\sunit$ (semantically, it \emph{merges}
capabilities provided by the declaration-site and call-site contexts). In structural systems such
as bounded reuse, $\apar$ is not symmetric and so this gives us a coeffect associated with the bound variable.

The (\emph{app}) rule follows the patterns seen
earlier -- it uses the scalar-vector multiplication ($\cclrd{t}\,\aseq\,\aclrd{S}$) on 
coeffects associated with $\Gamma_2$.
Using the syntactic reading, it then \emph{merges} context requirements for $\Gamma_1$ and $\Gamma_2$.
In the dual semantic reading,
it \emph{splits} the provided context into two parts passed to the sub-expressions.

The typing of let-binding (\emph{let}) corresponds to the typing of an expression $(\lambda x.e_2)~e_1$.
Syntactically, the context requirements are first split using $\azip$ and then re-combined
using $\apar$.

\paragraph{Structural rules.}
The coeffect-annotated context can be transformed using structural rules that are
not syntax-directed. These are captured by (\emph{ctx}), which uses a helper judgment representing
context transformations $\ctxtrans{ \coctx{\Gamma}{R} }{ \coctx{\Gamma'}{R'} }{\theta}$.
The rule models that a context used in the rule conclusion $\coctx{\Gamma'}{R'}$ can be transformed
to a context required by the assumptions $\coctx{\Gamma}{R}$ (using the semantic bottom-up reading).
In the rule, $\theta$ is a variable substitution generated by the transformation, which is 
used in the (\emph{contr}) rule.

Exchange and contraction decompose and reconstruct coeffect annotations 
using $\azip_{\sclrd{m},\sclrd{n}}$ (in assumption) and $\apar_{\sclrd{m},\sclrd{n}}$ (in conclusion). 
The shape subscripts are omitted, but we require the shapes to match using $\sclrd{m} = \slen{\Gamma_1}$
and $\sclrd{n} = \slen{\Gamma_2}$.

The (\emph{weak}) rule drops an ignored variable annotated with $\aweak$ (compare with (\emph{var})
annotated using $\aunit$). The (\emph{exch}) rule switches the values 
while (\emph{contr}) combines them using $\cpar$ to represent sharing of the context.
Finally, (\emph{sub}) represents sub-coeffecting that can be applied (point-wise) to any
individual coeffect.

@@

The coeffect system uses a general notion of context shape, but
it has been designed with structural and flat systems in
mind. The structural system is new in this paper and so we
look at it first.

Recall the coeffect shapes that characterise structural systems: the shape is formed by natural
numbers (with addition) modelling the number of variables in the
context. The coeffect algebra is therefore formed by the free monoid
(vectors) over a coeffect scalar. This means that the system keeps a vector
of basic coeffect annotations -- one for each variable. An empty
context (\eg{}, in the (\emph{const}) rule) is annotated with an empty
vector.

\begin{definition}
Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ a \emph{structural
coeffect system} has: 
\begin{itemize}{}
\item[--] Coeffect shape $\sclrd{(\mathbb{N}, {|}\textnormal{--}{|}, +, 0, 1)}$ formed by natural numbers
\item[--] Coeffect algebra $\aclrd{(\times, \times, \epsilon)}$ where $\times$ and $\epsilon$ are
  shape-indexed versions of the binary operation and the unit of a free monoid over $\C$.
  That is $\times : \C^{\sclrd{n}} \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{n + m}}$ appends vectors
  (lists) and $\epsilon : \C^{\sclrd{0}}$ represents empty vectors (lists)
\end{itemize}
\end{definition}

@@

\paragraph{Examples.} 
Defining a concrete structural coeffect system is easy, we just provide the coeffect scalar structure and the rest is free.

\begin{itemize}
\item
To recreate the system for bounded reuse, we use coeffect scalars formed by
$(\mathbb{N}, \ast, +, 1, 0, \leq)$. As in the system of Figure~\ref{fig:bounded-coeff},
\emph{used} variables are annotated with $1$ and \emph{unused} with $0$. Contraction adds the number
of uses via $+$ and application (sequencing) multiplies the uses.

\item
\emph{Dataflow} uses natural numbers (of past values), but differently: $(\mathbb{N}, +, \mathsf{max}, 0, 0, \leq)$.
Variables are initially annotated with $0$ (and can be incremented using the \kvd{prev} keyword).
Annotations of a shared variable are combined by taking maximum (of past values needed) and
sequencing uses $+$.

\item
Another use of the system is to track \emph{variable liveness}. The annotations are formed by
$\C=\{ \mathsf{D}, \mathsf{L} \}$ where $\mathsf{L}$ represents a \emph{live} (used) variable
and $\mathsf{D}$ represents a \emph{dead} (unused) variable. The scalars are given as
$(\C, \sqcap, \sqcup, \mathsf{L}, \mathsf{D}, \sqsubseteq)$.

In sequential composition ($\sqcap$), a variable is live only if it is required by both of the 
computations ($\mathsf{L} \sqcap \mathsf{L} = \mathsf{L}$), otherwise it is marked as 
dead ($\mathsf{D}$). A computation is not evaluated if its result is not needed.
A shared variable ($\sqcup$) is live if either of the uses is live 
($\mathsf{D} \sqcup \mathsf{D} = \mathsf{D}$, otherwise $\mathsf{L}$).
\end{itemize}

\noindent
Structural liveness is a practically useful, precise version of an example from our earlier
work, which was a flat system overapproximating liveness to the entire
context~\cite{petricek2013coeffects}.

@@

The same general coeffect system can be used to define systems that track whole-context coeffects
as in the implicit parameters example 
(Section~\ref{sec:intro-coeffects-impl}). Flat coeffect systems are
characterised by a singleton set of shapes, such as ${ \{ \sflat \} }$. In
this setting, the context annotations $\C^{ \sflat }$ are
equivalent to coeffect scalars $\C$.

In addition to the coeffect scalar structure, we also need to define $\apar$ and $\azip$.
Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\apar$ (merging of shaped 
coeffect annotations). However, the $\azip$ operation needs to be provided explicitly.

\begin{definition}
Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a binary 
$\cclrd{\wedge} : \C \times \C \rightarrow \C$ such that $(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$, we define:
\begin{itemize}{}
\item[--] Flat coeffect shape $\sclrd{(\{ \sflat \}, \emph{const} \, \sflat, \diamond, \sflat, \sflat)}$ where 
  $\sflat \diamond \sflat = \sflat$
\item[--] Flat coeffect algebra $(\cclrd{\wedge}, \cpar, \czero)$, \ie{}, the $\apar = \cpar$ and $\azero = 
\czero$ with the additional binary operation $\azip = \wedge$. 
\end{itemize}
\end{definition}

\noindent
The requirement $(\cclrd{r \wedge s}) \;\cleq\; (\cclrd{r}\,\cpar\,\cclrd{s})$ guarantees 
exchange and contraction preserve the coeffect of the assumption in the conclusion. Thus, flat coeffect
calculi do not require substructural-style rules.

\paragraph{Examples.}
Implicit parameters are the prime example of a flat coeffect system, but other examples
include rebindable resources \cite{app-distributed-acute} and Haskell type classes \cite{orchard2013thesis}.

In the \emph{implicit parameters} system (Section~\ref{sec:intro-coeffects-impl}), 
the co\-effect scalars are sets of names with types 
$\C = \mathcal{P}(\ident{Name} \times \ident{Types}) $. Variables
are always annotated with $\emptyset$ and coeffects are combined or split using set union $\cup$. Thus
the system is given by coeffect scalar structure $(\mathcal{P}(\ident{Name} \times \ident{Types}), \cup, \cup, \emptyset, \emptyset, \subseteq)$
with $\cclrd{\wedge} = \cup$.

\begin{remark}
We previously described flat systems for \emph{liveness} and \emph{dataflow}~\cite{petricek2013coeffects}.
Turning a structural system to flat requires finding $\wedge$ that 
underapproximates the capabilities of combined contexts. For dataflow, this is given by the \textsf{min} function as
$\mathsf{min}(\cclrd{r},\cclrd{s}) \,\cleq\, \mathsf{max}(\cclrd{r},\cclrd{s})$.

>>>>

\begin{itemize}
\item Structural coeffect calculus contains explicit rules for context manipulation
  (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not defined
  explicitly, but are admissible.

\item In the structural coeffect calculus, the variable context is treated as a vector
  and is annotated with a vector of (scalar) coeffects. In the flat coeffect calculus,
  the variable context is a set and is annotated with a single (scalar) coeffect.

\item In the flat coeffect calculus, we distinguish between splitting of the context requirements
  and merging of context requirements ($\cpar$ and $\czip$, respectively). In the structural
  coeffect calculus, the operations (which model splitting and appending vectors) are invertible 
  and so the structural coeffect algebra requires just $\atimes$.
\end{itemize}

@@

\subsection{Structure of coeffects}
In the structural coeffect calculus, the structure annotation was formed by a vector of coeffect
sclars. The unified coeffect calculus is similar, but a \emph{vector} is replaced with a 
\emph{container}. The primitive coeffect annotations in the unified calculus are formed by
a \emph{coeffect scalar}, which remains the same as in structural coeffect calculus
(Definition~\ref{def:structural-scalar}). In this section, we refer to it as
\emph{\cclrd{unified coeffect scalar}} (and we repeat the definition below). Then we
define \emph{\sclrd{unified coeffect containers}} which determines how coeffect scalar values 
are attached to the free-variable context. Finally, we define the \emph{\aclrd{unified coeffect 
algebra}} which consists of shape-indexed coeffect scalar values.

@@

\begin{example}
\label{ex:unified-struct-shape}
Structural coeffect container is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
where $\SH=\mathbb{N}$ and $\SP \sclrd{n}=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the 
number of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$, 
respectively, and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} = 
{|}\Gamma_1{|} + {|}\Gamma_2{|}$.

@@

\begin{example}
\label{ex:unified-flat-shape}
Flat coeffect container is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP \sflat = \{ 0 \}$
with a constant function ${|}\Gamma{|}=\sflat$ and a trivial operation $\sflat\,\diamond\,\sflat = \sflat$.

@@

\begin{definition}
Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a 
\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \emph{\aclrd{unified 
coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP \sempty}$
is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that 
combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
%
\begin{equation*}
\apar_{\sclrd{m},\sclrd{n}}, \azip_{\sclrd{m},\sclrd{n}} ~:~ \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
\end{equation*}
\end{definition}

\noindent
A coeffect algebra induces the following three additional operations:
%
\begin{equation*}
\begin{array}{rclcrcl}
 \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP \sunit}  \\
 \alift{\cclrd{x}} &\narrow{=}& \lambda \_ . \cclrd{x}   \\
\\[-0.5em]
 \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP \sclrd{m}} \rightarrow \C^{\SP \sclrd{m}}   \\
 \cclrd{r}\, \aseq\, \aclrd{\textbf{s}} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\, \cseq\, \aclrd{\textbf{s}}(\sclrd{s}) \\
\\[-0.5em]
 \alen{-} &\narrow{:}& \C^{\SP \sclrd{m}} \rightarrow \sclrd{m}\\
 \alen{\aclrd{\textbf{r}}} &\narrow{=}& \sclrd{m}
\end{array}
\end{equation*}
%
The $\alift{-}$ operation lifts a scalar coeffect to a shape-indexed coeffect that is indexed by 
the singleton context shape. The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a 
vector by scalar. As we always use bold font for vectors and ordinary font for scalars (as well 
as distinct colour), using the same symbol is not ambiguous. We also tend to omit the subscript
$\sclrd{m}$ and write $\aseq$.

@@

\vspace{1.5em}
{\small b.) Structural rules for context manipulation:}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s'}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{s}} \apar \aclrd{\textbf{q}}} \vdash e : \tau}
\quad
(\cclrd{s'} \cleq \cclrd{s})
\end{equation*}
\begin{equation*}
\tyrule{weak}
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \apar \alift{ \czero }} \vdash e : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s}} \azip \alift{\cclrd{t}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{t}} \apar \alift{\cclrd{s}} \apar \aclrd{\textbf{q}}} \vdash e : \tau}
\quad
\begin{array}{l}
 \slen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
 \slen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
\end{array}
\end{equation*}
\begin{equation*}
\hspace{2.5em} 
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s}} \azip \alift{\cclrd{t}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \apar \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
~
\begin{array}{l}
 \slen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
 \slen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
\end{array}
\end{equation*}

\vspace{0.5em}
\caption{Type system for the unified coeffect calculus}
\label{fig:unif-types}
\end{figure}

@@

\subsection{Unified coeffect type system}
\label{sec:unified-types}
The unified coeffect system in Figure~\ref{fig:unif-types} resembles the structural type system
shown in Figure~\ref{fig:struct-types}. Rather than explaining the rules one-by-one, we focus on
the key differences between the two.

The type system for the unified coeffect calculus is parameterized by a coeffect scalar 
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ together with a coeffect algebra $(\azip, \apar, \azero)$ 
and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$. 
As in the structural system, free-variable contexts $\Gamma$ are treated as vectors modulo duplicate
use of variables -- the associativity is built-int. The order of variables matters, but can be changed 
using the structural rules. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}} \in \C^\SH$ 
are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
annotated with coeffects scalars. % $\cclrd{r}, \cclrd{s}, \cclrd{t} \in \C$.

\paragraph{Syntax-driven rules.} 
The (\emph{var}) rule is syntactically the same as in the structural system, but it should be read
differently. The $\alift{-}$ operation does not create a \emph{vector}, but a \emph{shape-indexed
coeffect} that returns $\cunit$ for all positions in the singleton shape $\sunit$. 
The (\emph{const}) rule annotates empty context with a special annotation of the unit shape.

In a structural system, the two annotations correspond to a singleton and empty vector, respectively.
However, for a singleton shape with one position, the annotations are equivalent to annotating
variables with a scalar $\cunit$ and constants with a scalar $\czero$.

In the (\emph{app}), (\emph{abs}) and (\emph{let}) rules, the only change from the structural
system is that vector concatentation $\atimes$ is now replaced with context splitting/merging 
of the unified coeffect algebra. In particular, we use splitting of context requirements $\azip$
in rule \emph{assumptions} and merging of context requirements $\apar$ in rule \emph{conclusions}.

Note that we use the terms \emph{merging} and \emph{splitting} in the syntactic (top-down) sense. As 
discussed in Chapter~\ref{ch:flat}, Section~\ref{sec:flat-calculus-undestanding}, we can also read 
the rules in semantic (bottom-up) sense, in which case assumptions merge available contextual 
information and conclusions split available contextual information.

\paragraph{Structural rules.}
The merging/splitting operations in the structural rules are changed in the same way as in the
rules above. It is also worth noting that structural coeffect system used vectors of multiple 
elements. For example, $\alift{\cclrd{s},\cclrd{t}}$ and $\alift{\cclrd{t},\cclrd{s}}$ in the (\emph{exch}) 
rule denoted a two-element vector. In the unified system, this is replaced with merging/splitting
of two lifted scalars: $\alift{\cclrd{s}} \azip \alift{\cclrd{t}}$ and $\alift{\cclrd{t}} \apar \alift{\cclrd{s}}$.

In structural systems, this means the same thing -- we are simply concatenating or splitting two 
singleton vectors. However, the generalization allows us to capture flat coeffect systems as well.
The lifting operation in flat systems simply returns the lifted scalar and operators $\azip$ and
$\apar$ correspond to operations on coeffect scalars. And, as discussed in Section~\ref{sec:unified-flat},
thanks to the properties of coeffect scalars, contraction, weakening and exchange that do not 
affect the coeffect annotation are admissible for all flat systems.

@@

The unified coeffect system uses a general notion of context shape, but it has been designed with 
structural and flat systems in mind. In this and the next section, we show how it captures 
the two kinds of systems.

The unified calculus is modelled after the structural system and so using it to model structural
systems is easy -- given a coeffect scalar, we use the coeffect container that describes a \emph{vector} 
of annotations (Example~\ref{ex:unified-struct-shape}) and define a coeffect algebra formed by
a vector (free monoid) of scalars.

\begin{definition}
\label{def:unified-struct}
Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ a \emph{structural coeffect system} 
is defined by:

\begin{itemize}{}
\item[--] Coeffect container $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$ where $\SH=\mathbb{N}$ and
  $\SP \sclrd{n}=\{ 1 \ldots n \}$ and  ${|}x_1\!:\!\tau_1,\ldots,x_n\!:\!\tau_n{|}=n$.

\item[--] Coeffect algebra $\aclrd{(\times, \times, \epsilon)}$ where $\times$ and $\epsilon$ are
  shape-indexed versions of the binary operation and the unit of a free monoid over $\C$.
  That is $\times : \C^{\SP\sclrd{n}} \times \C^{\SP\sclrd{m}} \rightarrow \C^{\SP(\sclrd{n + m})}$ 
  appends vectors and $\epsilon : \C^{\SP\sclrd{0}}$ represents empty vectors.
\end{itemize}
\end{definition}

@@

\paragraph{Properties.}
An important property of the unified system is that, when used in a structural way as discussed
above, it gives calculi with the same properties as the structural system described in 
Chapter~\ref{ch:structural}. This can be easily seen by comparing the Figure~\ref{fig:unif-types}
with the Figure~\ref{fig:struct-types} and using the free monoid interpretation of the unified
coeffect algebra.

\begin{remark}
The system described in Definition~\ref{def:unified-struct} is equivalent to the structural coeffect
system described in Figure~\ref{fig:struct-types}. That is, a typing derivation using a structural
coeffect embedded in the unified system is valid if and only if the corresponding derivation is
valid in the structural system.
\end{remark}

\noindent
Using the above definition, our unified coeffect system can capture per-variable coeffect properties
discussed in Chapter~\ref{ch:applications}, Section~\ref{sec:applications-structural}. This includes
the system for bounded reuse (which is only used in the structural form) and precise tracking 
of per-variable dataflow and liveness.

@@

The same unified coeffect system can be used to capture systems that track whole-context (flat) 
coeffects such as implicit parameters. This is achieved using a singleton-shaped container for
coeffect annotations. The resulting system is has explicit structural rules (and is syntactically
different from the standard flat coeffect system), but we show that they are equivalent.

Flat coeffect systems are characterised by a singleton set of shapes (Example~\ref{ex:unified-flat-shape})
In this setting, the context annotations $\C^{ \SP \sflat } = \{0\} \rightarrow \C$ are equivalent 
to coeffect scalars $\C$. In addition to the coeffect scalar structure, we also need to define $\apar$ 
and $\azip$. Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\apar$ 
(merging of shaped coeffect annotations). The $\azip$ operation (corresponding to $\czip$
in flat coeffect calculus) needs to be provided explicitly.

\begin{definition}
\label{def:unified-flat}
Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a binary operation 
$\cclrd{\wedge} : \C \times \C \rightarrow \C$ such that $(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$, 
the unified coeffect algebra modelling flat coeffect systems consists of:
%
\begin{itemize}{}
\item[--] Flat coeffect container $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ as defined in 
  Example~\ref{ex:unified-flat-shape}.

\item[--] Flat coeffect algebra $(\czip, \cpar, \czero)$, \ie~$\apar = \cpar$ 
  and $\azero = \czero$ with an additional binary operation $\azip = \czip$. 
\end{itemize}
\end{definition}

\noindent
Intuitively, the requirement $(\cclrd{r}\,\czip\,\cclrd{s}) \;\cleq\; (\cclrd{r}\,\cpar\,\cclrd{s})$,
which could be also written as $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
denotes that splitting context requirements and then re-combining them preserves all the requirements
from the assumptions. The system may be imprecise and conclusions can overapproximate assumptions, but
it cannot lose requirements. This is fundamental for showing that exchange and contraction are admissible
in the unified system.

\paragraph{Properties.}
To show that the typing of flat properties in the unified system is equivalent to the typing in the
flat system, we show that a valid typing judgement in the first system is a valid typing judgement in 
the second system and vice versa.

In one direction, we show that the unified system (capturing flat properties) permits weakening, 
contraction and exchange rules that do not change the coeffect annotations. This guarantees that
a valid judgement in flat system is also valid in the unified system.

\begin{lemma}
\label{thm:unified-weak}
A unified coeffect calculus capturing flat properties admits weakening that does not change the
coeffect annotation.
\end{lemma}
\begin{proof}
The rule is admissible using the following derivation:
\begin{equation*}
\inference
  { \inference
      {\coctx{\Gamma}{ \cclrd{r} } \vdash e : \tau}
      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r} \;\apar\; \alift{ \czero }} \vdash e : \tau} }
  { \inference
      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r} \,\cpar\, \czero} \vdash e : \tau}      
      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r}} \vdash e : \tau} }
\end{equation*}

\noindent
We write $\cclrd{r}$ rather than $\aclrd{\textbf{r}}$, because we are tracking flat properties.
The first step is an application of the (\emph{weak}) rule. Next, we use the fact that
$\apar=\cpar$ and $\alift{\czero}=\czero$, which is the unit of the monoid $(\C, \cpar, \czero)$.
\end{proof}
